---
title: "ODE and forward sensitivity solving with `r2sundials`"
author: Serguei SOKOL
institute:  INRA / TBI / Mathematics cell
date: "December 12, 2019"
output:
  md_document:
    variant: gfm
---

<!-- to make Readme.md run
R --vanilla -e "rmarkdown::render('Readme.Rmd')"
-->
## Introduction
Package `r2sundials` is an RcppArmadillo wrapper for a well known and wide spread library [SUNDIALS/CVODES](https://computing.llnl.gov/projects/sundials/cvodes) from LLNL written in C. It provides an access from R to some basic features of `cvodes` module from this library which include:

 - solving real valued, user defined ODE via user provided functions calculating ODE's right hand side (rhs);
 - calculating first order forward sensitivities to parameters on which ODE solution depends ;
 - setting key parameters for ODE solving method like explicit or implicit time scheme, minimal or maximal steps, error order etc.
 
There is a copy of [SUNDIALS/CVODES](https://computing.llnl.gov/projects/sundials/cvodes) distributed with this package. So the package can be used just from its installation.

## Why another ODE solver for R?

The question is legitimate as there is already very furnished package [deSolve](https://cran.r-project.org/package=deSolve). It was also published [Rsundials](https://cran.r-project.org/package=Rsundials) package but it is archived since 2017-03-26. There is a fresh wrapper to the same library [sundialr](https://cran.r-project.org/package=sundialr). You can find even more packages dedicated to ODE solving in this [task view](https://CRAN.R-project.org/view=DifferentialEquations). So why a new wrapper?
Let see what are the novelties brought by `r2sundials` compared to most close alternatives: `deSolve` and `sundialr`.

### deSolve
Compared to `deSolve`, `r2sundials` provides possibilities:

 - to do forward sensitivity calculations for all or selected parameters;
 - to write users callback functions (rhs, Jacobian, ...) in RcppArmadillo where calculated values are stored "in-place" thus avoiding frequent memory reallocation;
 - to pass a parameter of any R type (vector, list, environment, ...) *directly* to Rcpp functions ;
 - to use [rmumps](https://cran.r-project.org/package=rmumps) package for solving underlying sparse linear systems;
 - to have much more flexible root finding and handling based on user callback functions.
 - to get statistics on ODE method used (call number for rhs routines, number of Jacobian calculations and so on).
 
### sundialr
Compared to `sundialr`, `r2sundials` provides:

 - more complete access to fine tuning of `cvodes` methods;
 - more complete parameter infrastructure. In `r2sundials`, parameters passed to callback functions can be of any R type (vector, list, environment, ...), not only numeric vector as in `sundialr`;
 - sensitivity calculations possibly done with the help of user provided functions (and not only with internal sundials procedure);
 - sensitivity calculations can be done on a selection of parameters, not necessarily on the totality of parameters.
 - Jacobian (dense or sparse) calculated with possibly user provided functions;
 - sparse system solving is made with `rmumps` package;
 - root finding and handling;
 - some statistics of ODE methods (call number for rhs routines, number of Jacobian calculations and so on).
 - (as of time of this writing, 2019-11-25) more thorough memory management which ensures that sundials' allocated memory is freed in due way, no matter what C++ exception and in what moment could happen. This avoids potential memory leaking problem in case of frequent package use during the same session.
 
## Install
The package can be installed as any other CRAN or github package:
```{r eval=FALSE}
install.packages("r2sundials")
```
or
```{r eval=FALSE}
devtools::install_github("sgsokol/r2sundials")
```

### Version note
`r2sundials` was developed and tested with CVODES version 5.0.0 released in October 2019. The versioning scheme of `r2sundials` is based on the underlying version of CVODES extended with one number proper to `r2sundials`. For example, `r2sundials` can have a version 5.0.0-2.

## Example
Let solve a very simple ODE $y'(t)=-ν(y(t)-a)$, with $ν=2$, $a=1$ and $y(0)=0$ on a time interval $[0, 3]$. This equation describes an exponential transition between two states 0 and $a$ with a rate $ν$.

With rhs written in R, we can do:
```{r}
library(r2sundials)
ti=seq(0, 3, length.out=101) # set time grid
p=c(nu=2, a=1) # set parameter vector
y0=0 # set initial condition
frhs=function(t, y, p, psens) -p["nu"]*(y-p["a"]) # set rhs functions
res=r2sundials::r2cvodes(y0, ti, frhs, param=p) # solve ODE
# compare with analytical solution
stopifnot(diff(range(p["a"]-exp(-p["nu"]*ti) - res)) < 1.e-6)
# see stats
print(attr(res, "stats"))
```

The same problem solved with RcppArmadillo rhs can look like:
```{r}
library(RcppXPtrUtils)
ptr_exp=cppXPtr(code='
int rhs_exp(double t, const vec &y, vec &ydot, RObject &param, NumericVector &psens) {
  NumericVector p(param);
  ydot[0] = -p["a"]*(y[0]-1);
  return(CV_SUCCESS);
}
', depends=c("RcppArmadillo","r2sundials","rmumps"),
 includes=c("// [[Rcpp::plugins(cpp14)]]", "using namespace arma;", "#include <r2sundials.h>"), 
 cacheDir="lib", verbose=FALSE)
# For ease of use in C++, we convert param to a numeric vector instead of a list.
pv=c(a=p$a)
# new call to r2cvodes() with XPtr pointer ptr_exp.
res2=r2sundials::r2cvodes(y0, ti, ptr_exp, param=pv)
stopifnot(diff(range(res2 - res)) < 1.e-14)
```

For more examples, see `?r2sundials::r2cvodes`.
